---
title: "Getting started with ZonationR"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval=FALSE, message=FALSE, warning=FALSE}
if (!require(devtools)) install.packages("devtools")
# Install the ZonationR package from GitHub
devtools::install_github("thiago-cav/ZonationR")

# Load libraries
library(ZonationR)
```

This vignette provides a brief introduction to **ZonationR** for beginners.
Users will:

1. Load example biodiversity data.
2. Prepare Zonation input files (feature list, settings, command file).
3. Run a single prioritization analysis.
4. Visualize the resulting prioritization map.

#### **Before you begin**

ZonationR is an interface to Zonation 5, which means that Zonation must be able to read and write files on disk. To ensure this works smoothly, it is recommended that you run the analysis inside an **RStudio project** whose directory is writable.

Before running the examples below, please make sure that:

- **Zonation 5 is installed** on your system  
- You know the **path to the Zonation 5 installation directory**  
- Your **current R project directory is writable**

All input and output files generated by *ZonationR* will be created within the current R project directory.

#### **Getting started**

This is a basic example to get a setup running with ZonationR. All files are created within the current R project, and the biodiversity input data used here are provided with the package.

```{r, eval = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# Load necessary libraries
library(ZonationR)
```

#### **Prepare input data**

First, we copy the example biodiversity data provided with the package into a local directory. These raster layers will be used as features in the Zonation analysis.

```{r, eval = FALSE, message = FALSE, results = 'hide', warning = FALSE}
#----------------------------------------------------
# Create a folder for biodiversity input data
#----------------------------------------------------
dir.create("biodiversity", showWarnings = FALSE)

extdata_path <- system.file("extdata", "biodiversity", package = "ZonationR")
file.copy(list.files(extdata_path, full.names = TRUE),
          "biodiversity", overwrite = TRUE)

```


#### **Zonation 5 setup**

Next, we create the three compulsory input files required by Zonation 5. The `zonation_path` argument must point to the directory where Zonation 5 is installed on your system.

```{r, eval = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# Create feature list file
feature_list(spp_file_dir = "biodiversity")

# Create settings file
settings_file(feature_list_file = "feature_list.txt")

# Create the Zonation command file
command_file(zonation_path = "C:/Program Files (x86)/Zonation5")

```

#### **Run the prioritization**

Finally, we run the Zonation 5 analysis. All outputs will be written to the project directory following the standard Zonation folder structure.

```{r, , eval = FALSE, message = FALSE, results = 'hide', warning = FALSE}

run_command_file(".")

```

#### **Keep Exploring**

Awesome! You just ran your first Zonation prioritization analysis 🎉 All outputs (priority map, performance curves, and other files) are in the `output` folder. Want to take it further? Run multiple scenarios, tweak feature weightings, or explore other settings in the ZonationR vignette...
