---
title: "Getting started with ZonationR"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette provides a brief introduction to *ZonationR*. Users will:

1. Set up example biodiversity data.
2. Prepare Zonation input files.
3. Run a single prioritization analysis.


> If you are new to spatial conservation prioritization or want a gentle overview to understand the basics, see the vignette [Why spatial conservation prioritization?](why_spatial_prioritization.html).


### Setup

```{r, message = FALSE, results = 'hide', warning = FALSE}
# Install the Zonation5RData package from GitHub:
pak::pak("thiago-cav/Zonation5RData")

# Load necessary libraries
library(ZonationR)
library(Zonation5RData)
```

### Before you begin

*ZonationR* is an interface to Zonation, which means that Zonation must be able to read and write files on disk. To ensure this works smoothly, it is recommended that you run the analysis inside an **RStudio project** whose directory is writable. Before running the example below, please make sure that:

- **Zonation 5 is installed** on your system  
- You know the **path to the Zonation 5 installation directory**  
- Your **current R project directory is writable**

To install the Zonation 5 software on your computer.

1. Visit the [official Zonation 5 website](https://zonationteam.github.io/Zonation5).

2. Download the appropriate installer or executable for your operating system (e.g., Windows, Linux).

3. Install or unpack the software, and note the installation path. You’ll need it later. 

### Prepare input data

For this example, we use the [Zonation5RData](https://github.com/thiago-cav/Zonation5RData) package, which provides GeoTIFF layers for workflows with Zonation and *ZonationR*. To use the biodiversity layers in the **RStudio project**, we first create a folder and then copy the raster files from the package.

```{r, eval = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# Create a folder for biodiversity input data
dir.create("biodiversity", showWarnings = FALSE)

# Get the paths to biodiversity rasters included in the package
files <- zonation5rdata_list("biodiversity", full.names = TRUE)

# Copy the raster files to the local biodiversity folder
file.copy(files, "biodiversity", overwrite = TRUE)
```


### Preflight checks

Now let’s check our setup! You can use the preflight functions in *ZonationR* to check that key requirements are met before creating input files or running Zonation. For example, you can quickly see if R can find the Zonation 5 executable on your computer. The installation path will be used later when creating the Zonation command file. 

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# Run one preflight check
z_check <- check_zonation_executable()

# Print the message
cat(z_check$message)

```

Zonation analyses require that input feature layers (e.g., species distribution rasters) share the same extent, resolution, and projection. The *ZonationR* package provides the function `check_raster_uniformity()` to verify that your layers are harmonized. You can also use `check_dir_writable()` to ensure that your working directory is writable.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# Run the other preflight checks
check_raster_uniformity("biodiversity")
check_dir_writable(".")

```

### Zonation input files

Next, we create the three compulsory input files required by Zonation 5. The `zonation_path` argument must point to the directory where Zonation 5 is installed on your system.

```{r, eval = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# Create feature list file
feature_list(spp_file_dir = "biodiversity")

# Create settings file
settings_file(feature_list_file = "feature_list.txt")

# Create the Zonation command file
command_file(zonation_path = "C:/Program Files (x86)/Zonation5")

```


### Run the prioritization

Finally, we run the Zonation 5 analysis. 

```{r, , eval = FALSE, message = FALSE, results = 'hide', warning = FALSE}
# Run the Zonation 5 analysis
run_command_file(".")

```

Awesome! You just ran your first Zonation prioritization analysis 🎉 All outputs (priority map, performance curves, and other files) are in the `output` folder. 


### Keep exploring

Want to take it further? Run multiple scenarios, tweak feature weightings, or explore other settings in the vignette [Working with variants in ZonationR](variants-workflow.html).<br><br>



