---
title: "Analysis flags in ZonationR"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


*ZonationR* provides a set of single-letter flags that control analysis options through `command_file()`. Certain flags require specific columns in the feature list (e.g., `weight` or `group`), while others need additional raster inputs supplied via `settings_file()`. As **additional material to the function documentation**, this vignette provides a guide to the available analysis flags and their relationships to feature and settings inputs, helping users better understand how to configure *ZonationR* workflows.

### Feature list

The `feature_list()` function produces a file specifying biodiversity features (e.g., species or habitat distributions). Users can assign weights, groups, and thresholds to control prioritization. Columns in the feature list, such as `weight`, `group`, and `threshold`, are used by the prioritization algorithm according to the flags set in `command_file()`.

```{r feature-list-table, echo=FALSE}
feature_list_table <- data.frame(
  Argument = c("filename", "weight", "group", "threshold"),
  ParameterFlags = c("-", "w", "g", "t"),
  DataType = c("string", "floating-point", "integer", "floating-point"),
  Explanation = c(
    "Feature distribution file name",
    "Feature weight",
    "Output group number",
    "Numeric threshold below which values are cut to zero"
  )
)

knitr::kable(
  feature_list_table,
  caption = "Table 1. Arguments accepted by `feature_list()` and their parameter flags"
)
```

### Settings file

The `settings_file()` function generates a configuration file that defines the inputs for a Zonation analysis. Some spatial inputs, such as analysis area masks or hierarchic masks, are optional but require activation via the corresponding flags in `command_file()`. File inputs include the feature list (i.e., the output from the `feature_list()` function), optional external solutions (`external solution`), analysis area masks (`analysis area mask layer`), hierarchic masks (`hierarchic mask layer`), and cost layers (`cost layer`).

```{r settings-file-table, echo=FALSE}
settings_file_table <- data.frame(
  FileParameter = c(
    "feature list file", "external solution", "analysis area mask layer",
    "hierarchic mask layer", "cost layer"
  ),
  CommandFlag = c("always required", "--mode=LOAD", "a", "h", "x, X"),
  RequiredFileType = c("plain text file", "floating-point raster", "integer raster", "integer raster", "floating-point raster")
)

knitr::kable(
  settings_file_table,
  caption = "Table 2. Arguments accepted by `settings_file()`"
)
```

### Command file

Analysis options in *ZonationR* are controlled via single-letter flags specified in `command_file()`. Some flags require corresponding columns in the feature list (e.g., `weight`, `group`, `threshold`), while others require additional raster inputs supplied through `settings_file()`. The table below summarizes the available flags, the required feature list columns, and any additional raster inputs needed.

```{r flags-table, echo=FALSE}
flags_table <- data.frame(
  Flag = c("a", "w", "g", "h", "x", "X", "t"),
  Explanation = c(
    "Use analysis area mask",
    "Apply feature weights",
    "Use output groups",
    "Hierarchic analysis",
    "Include cost in outputs only",
    "Include cost in prioritization",
    "Apply thresholding"
  ),
  FeatureListColumn = c("-", "weight", "group", "-", "-", "-", "threshold"),
  AdditionalRaster = c(
    "1 integer raster (analysis area)",
    "-",
    "-",
    "1 integer raster (hierarchic mask)",
    "1 floating-point raster",
    "1 floating-point raster",
    "-"
  )
)

knitr::kable(
  flags_table,
  caption = "Table 3. Analysis flags available in *ZonationR* and their requirements"
)
```
